---
title: Violin Plots
author: Stefania Ashby
date: '2019-12-09'
slug: violin-plots
categories: []
tags:
  - data visualization
  - violin plots
  - ggplot
subtitle: ''
summary: ''
authors: []
lastmod: '2019-12-09T23:28:22-08:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: vignette
    toc: true
    toc_float: false
    code_float: true
    df_print: paged
    number_section: true
---
```{r setup, include = FALSE}
# Clear the environment
rm(list = ls())

# Install and load required packages
list.of.packages <- c("tidyverse", "pander", "lme4", "rio", "lmerTest", "psych", "here", "janitor", "skimr", "magrittr", "cowplot", "wesanderson", "extrafont")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])] 
if (length(new.packages)) install.packages(new.packages, repos = "http://cran.us.r-project.org")
lapply(list.of.packages, library, character.only = TRUE)

# Turn off scientific notation
options(scipen = 999, digits = 4)
theme_set(theme_cowplot())


```


```{r import and tidy data, include = FALSE}
data <- import("~/Dropbox/data_science/facat_goodtrain_inc41.csv") %>% 
  select(subjNumber, mFirstName, mSurname, CHR, catACCold, catACCnew)

# Prepping the data for categorization plots
cat_plot_data <- data %>% 
  select(subjNumber, catACCold, catACCnew) %>% 
  gather(condition, accuracy, -subjNumber)

cat_plot_data$condition <- as.factor(cat_plot_data$condition)
str(cat_plot_data)

cat_labels = c("New", "Old")

# Prepping data for cued-recall plots
# Prepping the data
cr_plot_data <- data %>% 
  select(subjNumber, mFirstName, mSurname) %>% 
  gather(condition, perf, -subjNumber)

cr_plot_data$condition <- as.factor(cr_plot_data$condition)
str(cr_plot_data)

cr_labels = c("First", "Last")
```

# Intro
Violin plots allow us to look at the distribution of our data. 
But I know what you're thinking, "Can't I just use a density plot to do the same thing?".
While it's true you can use the density plot to show the same information, violin plots are better if you have multiple groups or conditions you need to plot in the same chart. 

Let me show you why:

# Density Plot
The example data I will use here comes from a manuscript that I am currently preparing for publication. Here I have two conditions that I'm plotting: 1. Categorization accuracy for old items, 2. Categorization accuracy for new items.

```{r density plot, warning = FALSE, message = FALSE}
pal = wes_palette("Darjeeling1", 2, type = "discrete") # Wes Anderson Palette's are fun! Check them out! 

## Build density plot
cat_density1 <- ggplot(cat_plot_data, aes(x = accuracy, fill = condition)) +
  geom_density() +
  scale_fill_manual(labels = cat_labels, values = pal) +
  labs(title = "Categorization",
       x = "Accuracy", 
       y = "Density") +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.justification = "center",
        plot.title = element_text(hjust = .5))
cat_density1

```

You can see that when we use a density plot, we get a nice look at the distribution of the two groups. However, they are overlapping. This may not be a big deal when we only have two conditions/groups we are comparing. But imagine how much more difficult this would be to visualize our data if we had 3 or more groups. 

One thing we could do is use the *facet_wrap* function to split our distributions into separate but side-by-side charts. 

```{r density plot 2, warning = FALSE, message = FALSE}

cat_density2 <- ggplot(cat_plot_data, aes(x = accuracy, fill = condition)) +
  geom_density() +
  scale_fill_manual(labels = cat_labels, values = pal) +
  labs(title = "Categorization",
       x = "Accuracy", 
       y = "Density") +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.justification = "center",
        plot.title = element_text(hjust = .5)) +
  facet_wrap(~condition, ncol = 1) +
  theme(strip.background = element_blank(), #Remove the condition labels since we have a legend
        strip.text.x = element_blank())
cat_density2

```

This looks pretty nice! But, violin plots allow us to look at the same information but with all groups included in the same chart. No duplicate y-axis! 

# Violin Plots
We can plot the same data on a single graph like so:

```{r basic violin plot, warning = FALSE, message = FALSE}
cat_v_basic <- ggplot(cat_plot_data, aes(x = condition, y = accuracy, fill=condition)) +
  geom_violin(trim = FALSE) +
  scale_fill_manual(values = pal) +
  ylab("Categorization Accuracy (% Correct)") +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 15),
        text = element_text(family = "Arial",
                            size = 25)) + 
  scale_y_continuous(breaks = c(0,.25, .5, .75, 1)) +
  scale_x_discrete(labels= cat_labels)
cat_v_basic

```

## Adding dots for individual differences
I can also superimpose individual dots for each subject to help visualize individual differences in the data.

```{r less basic violin plot, warning = FALSE, message = FALSE}
cat_l_basic <- ggplot(cat_plot_data, aes(x = condition, y = accuracy, fill=condition)) +
  geom_violin(trim = FALSE) +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = .75, fill = "black") + #added dots 
  scale_fill_manual(values = pal) +
  ylab("Categorization Accuracy (% Correct)") +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 15),
        text = element_text(family = "Arial",
                            size = 25)) + 
  scale_y_continuous(breaks = c(0,.25, .5, .75, 1)) +
  scale_x_discrete(labels= cat_labels)
cat_l_basic

```

## Adding mean and reference line
Want to know the average accuracy? I can also add a marker to denote the mean for each group and a reference line to show where chance performance lies (33% for three categories). I'll also space the dots further apart from one another so they're no longer touching. 

```{r fancy violin plot, warning = FALSE, message = FALSE}
pal = wes_palette("Darjeeling1", 2, type = "discrete")

## Build plot
cat_v_fancy <- ggplot(cat_plot_data, aes(x = condition, y = accuracy, fill=condition)) +
  geom_violin(trim = FALSE, scale = "count") +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = .75,stackratio = 1.5, fill = "black") +
  stat_summary(fun.y = mean, geom = "point", size = 3, shape = 23, fill = "Gold") + #adding mean marker
  scale_fill_manual(values = pal) +
  labs(title = "Categorization", 
       y = "Categorization Accuracy (% Correct)") +
  theme(legend.position = "none",
        plot.title = element_text(size = 20, hjust = .5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 15),
        text = element_text(family = "Arial",
                            size = 25)) + 
  scale_y_continuous(breaks = c(0,.25, .5, .75, 1)) +
  scale_x_discrete(labels= cat_labels) +
  geom_hline(yintercept = .333, linetype = "dashed", color = "black") + #added reference line
  ggsave("categorization_regdots.png", width = 2.6, height = 2.6, units = "in")
cat_v_fancy

```
